Notes on Validated Model Counting \\ Version of \today

### 2022-04-11 doc/mc-notes.tex
Update CRAT representation

\documentclass{llncs}
\usepackage{latexsym}
\usepackage{times}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{marvosym}
%% Colored hyperlink 
\newcommand{\cref}[2]{\href{#1}{\color{blue}#2}}
%% Colored hyperlink showing link in TT font
% \newcommand{\chref}[1]{\href{#1}{\small\tt \color{blue}#1}}
\newcommand{\hcref}[1]{\cref{#1}{\small\tt #1}}

@@

@@

\newcommand{\subs}[2]{[#2/#1]}
\newcommand{\substrue}[1]{\subs{#1}{\tautology}}
\newcommand{\subsfalse}[1]{\subs{#1}{\false}}
\newcommand{\subsflip}[1]{\subs{#1}{\obar{#1}}}

@@

For assignment $\interp$ and a Boolean formula $E$ over $X$, we
use the notation $\interp\subs{x_i}{E}$ to denote the assignment
$\interp'$, such that $\interp'(x_j) = \interp(x_j)$ for all $j \not = i$
and $\interp'(x_i) = \interp(E)$, where $\interp(E)$ is the value obtained by evaluating formula $E$ with each variable assigned the value given by $\interp$.
In particular, the notation
$\interp\subsflip{x_i}$ indicates the assignment in which the value
assigned to $x_i$ is complemented, while others remain unchanged.

A Boolean function $f$ is said to be {\em independent} of variable
$x_i$ if every $\interp \in \interpset{f}$ has $\interp\subsflip{x_i}
\in \interpset{f}$.  The {\em dependency set} of $f$, denoted
$\depend(f)$ consists of all variables $x_i$ for which $f$ is {\em
  not} independent.

@@

{\bf Example 1}: Let $\ring$ be the set of rational numbers and
$\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
assignment will be $1/2^{n}$, and the cost of a function will be its
{\em density}, denoted $\density(f)$.  That is $\density(f) \in [0,
  1]$ will be the fraction of assignments for which the function
evaluates to 1.  Given an ability to compute the density of
function $f$, we can scale this by $2^n$ to compute $|\interpset{f}|$, the core task of model counting.
Using density as our metric has the advantage that it
does not vary when the function is embedded in a larger domain $X'
\supseteq X$.  As a variant, we can support some forms of
weighted model counting by assigning weights of the form
$\cost(x_i) = w_i$, where $0 \leq w_i \leq 1$ for every variable $x_i$.

@@

Three key properties of separable cost functions makes it possible, in
some cases, to compute the cost of a Boolean formula without
enumerating all of its satisfying solutions.

@@

We introduce the CRAT clausal proof framework for generating a
checkable proof of the valuation of a Boolean formula, given in
conjunctive normal form, according to a specified cost function.  A
CRAT proof provides two capabilities: 1) a schema for computing a cost
function using the ring operations, and 2) a proof that this schema
accurately characterizes the cost function for the input formula.
A proof can be either {\em function independent}, meaning that it holds
for any cost function, or {\em function dependent}, meaning that
it holds for only for a specific cost function.

The CRAT format draws its inspiration from the QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item Clause addition is allowed, as long as either 1) the new clause is
  blocked, or 2) it satisfies the RAT property with respect to a sequence
  of earlier antecedent clauses.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and {\em abstraction}
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the proof step must cite earlier steps providing a RUP proof
  that the two arguments are mutually exclusive.
\item The {\em abstraction} operation allows an extension
  variable to serve as an alias for one of more other 
  variables.  Use of this rule renders the proof function dependent.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item Unlike DRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

A CRAT proof follows the same general form as a QRAT satisfaction
proof.  That is, starting with the set of input clauses it produces a
sequence of steps that both add and delete clauses.  At the final
step, there must be only a single unit clause consisting of some
variable or its complement.  Except for trivial cases, the final
variable will be an extension variable.  The sequence of extension
operations define the schema for computing the cost function.  That
is, a value for the cost function is computed for every literal
according to the rules of Lemmas
\ref{lemma:complementation}--\ref{lemma:disjunction}.  For every
abstraction step, all variables having the indicated
extension variable as their alias must evaluate to the same value.

@@

Consider Boolean formulas over a set of variables $X$.  An
assignment $\interp$ is a function mapping each variable to 
truth value $\tautology$ (true) or $\nil$ (false).  We can extend
$\interp$ to Boolean formulas in the normal way, such that
$\interp(F)$ will be $\tautology$ (respectively, $\nil$) if the
evaluation of formula $F$ yields $\tautology$ (resp., $\nil$) when its
variables are assigned values according to $\interp$.

For Boolean formula $F$, we define its set of {\em models} $\interpset{F}$ as
\begin{eqnarray}
\interpset{F} & = & \{ \interp | \interp(F) = \tautology \}
\end{eqnarray}
The task of model counting is, given formula $F$, to determine the size of its set of models
$\mcount(F) = |\interpset{F}|$.  Ideally, this should be done without actually enumerating the set.

For assignment $\interp$ and a Boolean formula $E$ over $X$, we
use the notation $\interp\subs{x}{E}$ to denote the assignment
$\interp'$, such that $\interp'(y) = \interp(y)$ for all $y \not = x$
and $\interp'(x) = \interp(E)$.  In particular, the notation
$\interp\subsflip{x}$ indicates the assignment in which the value
assigned to $x$ is complemented, while others remain unchanged.

A Boolean formula $F$ is said to be {\em independent} of variable $x$
if every $\interp \in \interpset{F}$ has $\interp\subsflip{x} \in
\interpset{F}$.

The following lemmas describe the primary abstractions used to perform
model counting without enumerating all solutions.

\begin{lemma}
  \label{lemma:complement}
  For any formula $F$:
  \begin{eqnarray*}
    \mcount(\neg F) & =  & 2^n - \mcount(F)
  \end{eqnarray*}
\end{lemma}

\begin{lemma}
  \label{lemma:product}
  For a partitioning of the variables $X$ into disjoint sets $X_1$ and $X_2$, and for formulas $F_1$ defined over $X_1$ and $F_2$ defined over $X_2$:
  \begin{eqnarray*}
    \mcount(F_1 \land F_2) &=& \mcount(F_1) \cdot \mcount(F_2)
  \end{eqnarray*}
\end{lemma}

\begin{lemma}
  \label{lemma:sum}
  For formulas $F_1$ and $F_2$ such that $\interpset{F_1} \cap \interpset{F_2} = \emptyset$:
  \begin{eqnarray*}
    \mcount(F_1 \lor F_2) &=& \mcount(F_1) + \mcount(F_2)
  \end{eqnarray*}
\end{lemma}

\section{ITE Operation}

We consider a single Boolean operation, known as
``If-Then-Else,'' or simply ``ITE.''  For Boolean values $a, b, c$,
the operation is defined as $\ite(a,b,c) = (a \land b) \lor (\neg a
\land c)$.  This single operation can be used to express several common Boolean operations:
\begin{eqnarray*}
\neg a & = & \ite(a, \nil, \tautology) \\
a \land b & = & \ite(a, b, \nil) \\
a \lor b & = & \ite(a, \tautology, b) \\
a \rightarrow b & = & \ite(a, b, \tautology) \\
\end{eqnarray*}

For variable $x$ and Boolean formulas $F$ and $G$, the set of models
for $\ite(x, F, G)$ is given by the formula:
\begin{eqnarray}
  \interpset{\ite(x, F, G)} & = & \{ \interp \in \interpset{F} | \interp(x) = \tautology \}
  \cup \{ \interp \in \interpset{G} | \interp(x) = \nil \} \label{eqn:ite:interp}
\end{eqnarray}

\section{ITE Graphs}

An {\em ITE graph} describes a combinational logic circuit constructed from two-input multiplexors.
It is defined as a directed acyclic graph
with three node types:
\begin{description}
\item[Constant:] Corresponds to value $\tautology$ or $\nil$.  A constant node has no incoming arcs.
\item[Input:] Corresponds to one of the variables in $X$.   A variable node has no incoming arcs.
\item[Operator:] Represents an application of the $\ite$ operation.  An operator node has three incoming arcs, labeled \ifarg{}, \thenarg{}, and \elsearg{}, corresponding to the three arguments of the $\ite$ operation.
\end{description}

When describing ITE graphs, we refer to constant and input
nodes by their associated value or variable.  An operator node is described by an expression of the form
$\ite(v_i, v_t, v_e)$, where $v_i$, $v_t$, and $v_e$ are the
nodes corresponding to the \ifarg{}, \thenarg{}, and \elsearg{} inputs, respectively.

We can define the function $\depend$ mapping each node
$v$ in an ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
\begin{eqnarray}
\depend(\tautology) & = & \emptyset \label{eqn:depend:true}\\
\depend(\nil) & = & \emptyset \label{eqn:depend:false}\\
\depend(x) & = & \{x \} \label{eqn:depend:variable}\\
\depend(\ite(x, v_t, v_e)) & = & \depend(v_i) \cup \depend(v_t) \cup \depend(v_e) \label{eqn:depend:ite}
\end{eqnarray}

\begin{lemma}
\label{lemma:independent:dset}
Any node $v$ in an ITE graph is independent of any variable $y \in X$ such that $y \not \in \depend(v)$.
\end{lemma}  
\noindent
{\em Proof:} Follows by the recursive definition of $\depend$ (\ref{eqn:depend:true}--\ref{eqn:depend:ite}) and by (\ref{eqn:ite:interp}).

A {\em variable-partitioned} ITE (VPITE) graph is one satisfying the property
that for every operator node $v = \ite(v_i, v_t, v_e)$, the dependency
set of its \ifarg{} input most be disjoint from those of its
\thenarg{} and \elsearg{} inputs.  That is, both $\depend(v_i) \cap \depend(v_t) = \emptyset$ and
$\depend(v_i) \cap \depend(v_e) = \emptyset$.

\section{Model Counting VPITE Graphs}

While model counting is a difficult problem for arbitrary
formulas, it readily be computed when $F$ takes the form of a VPITE graph.

\begin{theorem}
\label{theorem:count}
Letting $n = |X|$, the following recursive formula holds for any node in a VPITE graph:
\begin{eqnarray}
\mcount(\tautology) & = & 2^{n} \label{eqn:count:true}\\
\mcount(\nil) & = & 0 \label{eqn:count:false}\\
\mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2^n} \label{eqn:count:ite}
\end{eqnarray}  
\end{theorem}
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemmas~\ref{lemma:complement}--\ref{lemma:sum}, along with the definition of the ITE operator.

\section{Relation to Other Representations}

In general, there is a trade-off between the compactness of different
Boolean function representations versus the ease with which different
properties can readily be computed.  For example, Boolean formulas and
Boolean circuits can be very compact, but basic properties of these
representations, such as satisfiability, equivalence testing, and
model counting are NP-hard.  On the other hand all of these properties
can be computed in polynomial time for reduced, orderd binary decision
diagrams (ROBDDs)~\cite{Bryant:1986}, but they can grow exponentially
in relation to the representation of the function as a formula or
circuit.  ITE graphs lie somewhere between these two extremes.  Here
we examine how they relate to other representations and the implications
of these relations.

\subsection{Binary Decision Diagrams}

Binary Decision Diagrams~\cite{akers:ieeetc:1978} also represent Boolean formulas
as directed graphs consisting of ITE operators.  They can be seen as
equivalent to a class of graphs we will call {\em input-controlled} ITE
graphs.  These are ones for which the \ifarg{} input to every operator
node is a variable.

If we combine the properties of variable partitioned and input
control, we arrive at the representation known as {\em
  Free}~BDDs~\cite{gergov:ieeetc:1994}.  If we further restrict the
variables in a variable-controlled, input-partitioned ITE graph to
obey the same ordering along all paths formed in the graph, we arrive
at OBDDs.  An OBDD can be reduced by a simple set of rules to give an
ROBDD representation, and this form is canonical~\cite{Bryant:1986}.

\subsection{Sententential Decision Diagrams}

Darwiche uses the Sentential Decision Diagram (SDD)
representation~\cite{darwiche:ijcai:2011} as his framework for several
model counting programs.  An SDD can be seen as a combination of 1) a
generalization of VPITE graphs consisting of $k$-input multiplexors,
and 2) an ordering restriction on the variables.
That is, each operator node in an SDD is defined to
have $k$ data inputs and $k$ control inputs, where the Boolean
functions corresponding to the control inputs $p_1, p_2, \ldots, p_k$
satisfy $p_i \land p_j = \nil$ for $1 \leq i < j \leq k$ and
$\bigvee_{i=1,k} p_i = \tautology$.  The $k$ data inputs represent
Boolean functions $s_1, s_2, \ldots, s_k$.  In addition, the
dependency sets for the control and data inputs must be disjoint:
$\depend(p_i) \cap \depend(s_j) = \emptyset$ for $1 \leq i,j \leq k$.
Furthermore, the ordering of the variables in an SDD must obey a hierarchical decomposition
defined by
a tree structure known as a {\em vtree}.  Indeed, an SDD can be seen
as equivalent to a generalization of BDDs devised by McMillan, which
he termed {\em tree} BDDs~\cite{mcmillan:cav:1994}.

The graph part of an SDD can be readily translated into a VPITE graph
by expanding each $k$-input node into $k-1$ ITE nodes, giving an
output value defined as:
\begin{displaymath}
\ite(p_1, s_1, \ite(p_2, s_2, \ite( \cdots \ite(p_{k-1}, s_{k-1}, s_k)\cdots))
\end{displaymath}
If we express the size of an SDD as the number of branches, this
translation from SDD to ITE graph will expand the size by a constant
factor.

\section{Probabilistic Equivalence Checking}

One further feature of VPITE graphs is that there is a randomized
polynomial time (RP) algorithm for determining whether two graphs
represent equivalent Boolean functions.  This algorithm generalizes
the RP equivalence-testing algorithm for free BDDs described by Blum,
Chandra, and Wegman~\cite{blum:ipl:1980}.

\section{A Model Counting Calculus}

Objective: Create a simple system for defining and reasoning about an
abstraction of Boolean formulas that suffices for tracking the number
of models.

Define a {\em counting subspace} $\subspace{} = \langle V, \alpha, k\rangle$, with:
\begin{itemize}
\item $V$ is a set of variables
\item $\alpha$ is a partial assignment to variables in $V$, represented as a set of literals.
\item $k$ is an integer, indicating the number of models for
  the formula within the specified subspace.  For $n = |V|$ and $m= |\var(\alpha)|$, it satisfies
  $0 \leq k \leq 2^{n-m}$.
\end{itemize}

\subsection{Subspace Primitives}
\begin{itemize}
\item
Universal subspace: $\langle \emptyset, \emptyset, 1 \rangle$.
\item
Null subspace: $\langle \emptyset, \emptyset, 0 \rangle$.
\item
  Positive literal: $\langle \{ x\}, \{x\}, 1\rangle$.
\item
  Negative literal: $\langle \{ x\}, \{\obar{x}\}, 1\rangle$.
\item
  Unused variable: $\langle \{ x\}, \emptyset, 2\rangle$.
\end{itemize}

\subsection{Basic Subspace Operations}

Assume $\subspace_1 = \langle V_1, \alpha_1, k_1\rangle$ and
$\subspace_2 = \langle V_2, \alpha_2, k_2\rangle$ are two subspaces

Subspace product $\subspace_1 \times \subspace_2$
\begin{itemize}
\item Condition: $V_1 \cap V_2 = \emptyset$
\item Result: $\langle V_1 \cup V_2, \alpha_1 \cup \alpha_2, k_1 \cdot k_2\rangle$.
\end{itemize}

Subspace Sum $\subspace_1 + \subspace_2$
\begin{itemize}
\item Conditions:
  \begin{itemize}
  \item $V_1 = V_2$.
  \item For some literal $l$ of variable $x \in V_1$ and assignment $\alpha$, the argument assignments satisfy
    $\alpha_1 = \alpha \cup \{l\}$ and $\alpha_2 = \alpha \cup \{ \obar{l} \}$.
  \end{itemize}
\item Result: $\langle V_1 - \{ x\}, \alpha, (k_1 + k_2)/2\rangle$.
\end{itemize}

\subsection{Derived Subspace Operations}

These useful operations can be constructed by combining primitive values and basic operations.

Expanding subspace $\subspace{} = \langle V, \alpha, k\rangle$:
\begin{itemize}
\item
  Condition: $x \not \in V$
\item
  Results:
  \begin{itemize}
    \item Positive expansion: $\langle V \cup \{ x \}, \alpha \cup \{ x \}, k \rangle$.
    \item Negative expansion: $\langle V \cup \{ x \}, \alpha \cup \{ \obar{x} \}, k \rangle$.
    \item Don't-care expansion: $\langle V \cup \{ x \}, \alpha , 2\cdot k \rangle$.
  \end{itemize}
\end{itemize}

A particularly useful expansion is to apply the don't-care expansion
with a set of variables $U$ such that $U \cap V = \emptyset$.  This
will give the subspace $\langle U \cup V, \alpha, 2^{|U|} \cdot
k\rangle$.

>>>>

\documentclass{llncs}
\usepackage{latexsym}
\usepackage{times}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{cite}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{marvosym}
%% Colored hyperlink 
\newcommand{\cref}[2]{\href{#1}{\color{blue}#2}}
%% Colored hyperlink showing link in TT font
% \newcommand{\chref}[1]{\href{#1}{\small\tt \color{blue}#1}}
\newcommand{\hcref}[1]{\cref{#1}{\small\tt #1}}

@@

\newcommand{\makenode}[1]{{\mathbf #1}}
\newcommand{\nodeu}{\makenode{u}}

@@

\newcommand{\subs}[2]{[#2/#1]}
\newcommand{\substrue}[1]{\subs{#1}{\tautology}}
\newcommand{\subsfalse}[1]{\subs{#1}{\false}}

@@

For assignment $\interp$ and a Boolean formula $E$ over $X$, we
use the notation $\interp\subs{x_i}{E}$ to denote the assignment
$\interp'$, such that $\interp'(x_j) = \interp(x_j)$ for all $j \not = i$
and $\interp'(x_i) = \interp(E)$, where $\interp(E)$ is the value obtained by evaluating formula $E$ with each variable assigned the value given by $\interp$.
In particular, the notation
$\interp(\subs{x_i}{\obar{x}_i})$
indicates the assignment in which the value
assigned to $x_i$ is complemented, while others remain unchanged.

A Boolean function $f$ is said to be {\em independent} of variable
$x_i$ if every $\interp \in \interpset{f}$ has $\interp(\subs{x_i}{\obar{x}_i})
\in \interpset{f}$.  The {\em dependency set} of $f$, denoted
$\depend(f)$ consists of all variables $x_i$ for which $f$ is {\em
  not} independent.

@@

{\bf Example 1}: Let $\ring$ be the set of rational numbers and
$\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
assignment will be $1/2^{n}$, and the cost of a function will be its
{\em density}, denoted $\density(f)$.  That is $\density(f) \in [0,
  1]$ will be the fraction of assignments for which the function
evaluates to 1.  Given an ability to compute the density of function
$f$, we can scale this by $2^n$ to compute the total number of models
$|\interpset{f}|$.  This is the core task of model counting.  Using
density as our metric has the advantage that it does not vary when the
function is embedded in a larger domain $X' \supseteq X$.  As a
variant, we can support some forms of weighted model counting by
assigning weights of the form $\cost(x_i) = w_i$, where $0 \leq w_i
\leq 1$ for every variable $x_i$.

@@

Three key properties of separable cost functions make it possible, in
some cases, to compute the cost of a Boolean formula without
enumerating all of its satisfying solutions.

@@

We introduce the CRAT clausal proof framework for generating a
checkable proof of the valuation of a Boolean formula, given in
conjunctive normal form, according to a specified cost function.  A
CRAT proof provides two capabilities: 1) a {\em schema} for computing a cost
function using the ring operations, and 2) a {\em proof} that this schema
accurately characterizes the cost function for the input formula.
A schema can be either {\em function independent}, meaning that it holds
for any cost function, or {\em function dependent}, meaning that
it holds for only for a specific cost function.

The CRAT format draws its inspiration from the QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item Clause addition is allowed, as long as either 1) the new clause is
  blocked, or 2) it satisfies the RUP/RAT property with respect to a sequence
  of earlier antecedent clauses.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and {\em abstraction}
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the step must cite earlier steps providing a RUP/RAT proof
  that the two arguments are mutually exclusive.
\item The {\em abstraction} operation allows an extension
  variable to serve as an alias for one of more other 
  variables.  Use of this rule renders the schema function dependent.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item Unlike DRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

\subsection{Syntax}

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[7mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
    $C$ & {\tt ab} & $L^{+}$ {\tt 0} & ${\tt -}C^{*}$  {\tt 0}  & Add blocked clause \\
    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RUP/RAT clause \\
     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RUP/RAT clause \\
    \midrule
        & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
        & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
        & {\tt a} & $V \; L$         & $V^{+}$ {\tt 0}  & First application of abstraction \\ 
        & {\tt a} & $V \; L$         &                  & Subsequent applications of abstraction \\ 
  \end{tabular}
  }
\end{table}

Table~\ref{tab:crat:syntax} shows the set of proof rules for the CRAT
format.  As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $l$
is represented by a signed integer, with $-v$ being the complement of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being input clauses and successive
ones being added clauses.  Clause identifiers must be totally ordered,
such that clause $C$ can only reference clauses $C'$ such that $C' <
C$.  However, clause identifiers need not be consecutive.

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via
blocked-clause and RUP/RAT rules.  The hints portion of a
blocked-clause addition lists all earlier clauses containing
the negated version of the pivot literal, with the clause IDs negated.  The hints portion of a
resolution addition must contain a sequence of clause IDs such that
the added clause is RUP/RAT with respect to these clauses.  Clause
deletion requires an explicit justification that the deleted clauses
is RUP/RAT with respect to other clauses.

The second set of proof rules is unique to the CRAT format.  Each of
these indicates the addition of an extension variable.  For each case,
the rule must be followed by a sequence of blocked-clause additions.
A product rule of the form ${\tt p}\;v\;l_1\;l_2$ indicates that $v$
will represent the product $l_1 \pand l_2$.  The blocked clause
additions must encode the formula $v \leftrightarrow (l_1 \land l_2)$.
Literals $l_1$ and $l_2$ must have disjoint dependency sets.

A sum rule of the form ${\tt s}\;v\;l_1\;l_2$ indicates that $v$ will
represent the disjunction $l_1 \por l_2$.  The blocked clause
additions must encode the formula $v \leftrightarrow (l_1 \lor l_2)$.
The rule also contains a sequence of clause IDs such that the clause
$\obar{l}_1 \lor \obar{l}_2$ is RUP/RAT with respect to the sequence.

An abstraction rule of the form ${\tt a}\;v\;l$ indicates that
variable $v$ will be an alias for literal $l$.  The syntax differs
according to whether or not this is the first use of $v$ as an
alias.  The first instance must include a list of variables that is a
(possibly improper) superset of the dependency set for $l$.  It must
be followed by a sequence of blocked-clause additions encoding the
formula $v \leftrightarrow l$.  Subsequent instances do not
include an explicit dependency set, but $\depend(l)$ must be a subset of
those declared for $v$.  Furthermore, there are no blocked-clause
additions for these cases.

\subsection{Semantics}

A CRAT proof follows the same general form as a QRAT dual proof---one
that ensures that each clause addition and each clause deletion preserves
equivalence.  Starting with the set of input clauses, it produces a
sequence of steps that both add and delete clauses.  Each addition
must be truth preserving and each deletion must be falsehood
prserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause consisting of some variable or
its complement.  Except for trivial cases, the final variable will be
an extension variable.  The sequence of extension operations define
the schema for computing the cost function.  That is, a value for the
cost function is computed for every literal according to the rules of
Lemmas \ref{lemma:complementation}--\ref{lemma:disjunction}.  For
every abstraction step, all literals having the indicated extension
variable as their alias must evaluate to the same value.  The computed
cost is then the value computed for the literal in the final unit
clause.

\subsection{Example}

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
    A) Logical Schema\\[2ex]
    \centering{\input{dd/c3}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Evaluation Schema\\[2ex]
    \centering{\input{dd/c3-eval}}
    \end{minipage}
  }
\caption{Schema for Formula $x_1 \lor x_2 \lor x_3$}
\label{fig:c3:schema}
\end{figure}

\begin{figure}
  \centering{
  \begin{tabular}{llllll}
    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
\midrule
    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause}\\
            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
    {\tt 2} & {\tt ab}  & {\tt 4  2 -3 0}  & {\tt 0}    & & Defining clauses for $p_4$ \\ 
    {\tt 3} & {\tt ab}  & {\tt -4 -2 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
    {\tt 4} & {\tt ab}  & {\tt -4 3 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
            & {\tt s}   & {\tt 5 2 4}   & {\tt 3 0}  & & Declare $s_5 = x_2 \por p_4$ \\
    {\tt 5} & {\tt ab}  & {\tt -5 2 4 0}  & {\tt 0}    & & Defining clauses for $s_5$ \\ 
    {\tt 6} & {\tt ab}  & {\tt  5 -2 0}    & {\tt -5 0} &  \\  %%Defining clause for $s_5$ \\ 
    {\tt 7} & {\tt ab}  & {\tt  5 -4 0}    & {\tt -5 0} & &  \\  %%Defining clause for $s_5$ \\ 
            & {\tt p}   & {\tt 6 -1 5}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
    {\tt 8} & {\tt ab}  & {\tt 6  1 -5 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
    {\tt 9} & {\tt ab}  & {\tt -6 -1 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
    {\tt 10} & {\tt ab}  & {\tt -6 5 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
             & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
    {\tt 11} & {\tt ab}  & {\tt -7 1 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
    {\tt 12} & {\tt ab}  & {\tt  7 -1 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 13} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
    {\tt 14} & {\tt ar}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
             & {\tt dr}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
  \end{tabular}
  }
  \caption{CRAT Proof Steps for Formula $x_1 \lor x_2 \lor x_3$}
  \label{fig:c3:crat}
\end{figure}

As an illustration, consider the Boolean formula $x_1 \lor x_2 \lor
x_3$, represented by a single clause.  We cannot directly use the
$\por$ operation to form this disjunction, since the sets of assignments
satisfying the individual literals are not disjoint.  Instead, we must
decompose this formula into a sequence of operations, as shown by the
logical schema of Figure~\ref{fig:c3:schema}A\@.  Circles in this schema
correspond to $\pand$ operations and squares to $\por$ operations.
The subscripts of the variables and the operator labels correspond to
the numbers of the input and extension variables in the CRAT proof.

The conjunction of $\obar{x}_2$ and $x_3$ can be computed as $p_4 =
\obar{x}_2 \pand x_3$, since the literals have disjoint dependency
sets. We can then express the disjunction $x_2 \lor x_3$ as
as $s_5 = x_2 \por p_4$.  A similar process forms the
disjunction $x_1 \lor x_2 \lor x_3$ by first forming the product $p_6
= \obar{x}_1 \pand s_5$ and the final sum $s_7 = x_1 \por p_6$.

The logical representation can readily be converted into a schema for
computing the cost of the formula, given weight $w_i$ for each
variable $x_i$ for $1 \leq i \leq 3$.  This is illustrated in
Figure~\ref{fig:c3:schema}B\@.  This schema is valid for any cost
function, since it contains no abstraction operations.

Figure~\ref{fig:c3:crat} shows an annotated version of the CRAT proof
for this example.  Clause \#1 corresponds to the input formula, and
clauses \#2--\#13 are the defining clauses for the four operations.
Clause \#13 adds the unit clause corresponding to sum $s_7$,
indicating that the extension variable variable will evaluate to
$\tautology$.  for any assigment that satisfies the input clause.  We
can write this as $C \turnstile  s_7$ for input clause $C$.  The
deletion step at the end turns this around, showing that
$s_7 \turnstile C_I$, and therefore the input clause can be deleted,
This completes a proof that extension variable $s_7$ is logically
equivalent to the input formula.

\section{What's known, and what's not}

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.  The key idea of that work
was to use extended resolution to encode the semantics of the BDD
nodes as part of the proof.  Here we can further decompose each $\ite$
(short for ``if-then-else'') operation representing a BDD node into
two $\pand$ and one $\por$ operation.  That is, we can encode
$\ite(x,A,B$ as $(x \pand A) \por (\obar{x} \pand B$, where $x$ is an
input variable and $A$ and $B$ are subformulas.  In doing so, we must
make sure that $x$ is not in the dependency set of either $A$ or $B$.
For ordered BDDs,~\cite{Bryant:1986} this property holds, since the
variables in $A$ and $B$ must be greater in the variable ordering than
$x$.

The BDD representations of many of the formulas occuring in
model-counting problems are far to large for this approach to be
practical.  One refinement of the approach, implemented by the ADDMC
model counter, is to abstract subformulas, keeping track only of the
number of models they can have, representing a number of subformulas
with a single ADD leaf node.  Our abstraction operation is intended to
support this capability.  Under controlled conditions, we can
introduce a new variable the represents the cost function for some
subformula.  If subsequent formulas are guaranteed to yield the same
cost, then these can alias to the earlier variable.  However, it is
not clear how to prove that this aliasing preserves equivalence.

Other model counters use either top-down and bottom-up methods to
generate representations of the formula that are similar to our
schmas.  These exploit the key abstractions we have identified.  We
must find ways to modify these model counters to also generate CRAT
proofs.

@@
